
\chapter{Language design}
\label{chap3:language_design}

When creating Giflang we had two options:
\begin{enumerate}
    \item Use a~language with an~existing web interpreter
    \item Create a~new language and a~web interpreter
\end{enumerate}

In~the~first section of this chapter we explain~why we decided to not reuse an~existing language. In this section we discuss the~design of our own language. We do not
precisely define all the~semantics, but rather provide a~high-level overview. We think that it is more important to explain~\emph{why} Giflang uses specific
constructs, rather than~list \emph{how} these constructs behave in~every case.

The~language documentation with examples can~be found in~the~web IDE by clicking ``Docs'' button in the top navigation.

\section{Reusing an~existing interpreter}
in~the~previous Chapter \ref{chap2} we concluded that interpreting a~user code on the~client-side is the~best option for our project. Therefore, when
looking for a~language to reuse we need to limit ourselves to the~ones that have an~existing web interpreter, i.e., an~interpreter written in~either
JavaScript or WebAssembly. The~two major languages that have existing web interpreters are Python and JavaScript.

\subsection{Python}
There are multiple Python web interpreters, for example Pyodide \cite{Pyodide} or Brython \cite{Brython}. Pyodide works by compiling CPython \cite{CPython}
to WebAssembly using Emscripten. Brython works by transpiling Python to JavaScript. We were not able to find a~Python web interpreter that would
support code-stepping out of the~box. We considered forking either Pyodide or Brython and adding this functionality.

Adding code-stepping to Pyodide would require changing the~underlying CPy\-thon implementation. CPython is very complex and was not built for this use case.
On the~other hand, Brython is fully written in~JavaScript and also builds an~AST from a~Python code in~JavaScript. There
is even a~fork of Brython \cite{PythonDebugger} that implements code-stepping. However, at closer inspection we found out it runs the~code all at once and records trace information
at each executed line by inserting a~trace function call there. Afterwards, it only gives users an~illusion of code-stepping.

It is possible to create a~``live'' code-stepping for Brython in~a~similar fashion, for example with the~aid of Atomics API discussed in~Section \ref{chap4:atomics}.
We did not find a~simple way of increasing the~stepping granularity from just lines to arbitrary AST nodes, though. We definitely could have adjusted Brython instead
of creating a~custom interpreter. However, it would take a~non-trivial effort to change the~Brython's internals and we rather put this effort into creating
a~new interpreter from scratch and learning a~lot about languages and interpreters along the~way. Additionally, we did not have a lot of experience with writing
interpreters and compilers prior to this project and hence it would be quite challenging to customize an existing one.

\subsection{JavaScript}
\label{chap3:javascript}
There are numerous JavaScript interpreters written in~JavaScript. The~one closest to our needs is JS-Interpreter \cite{JSInterpreter} that supports running JavaScript
with code-stepping in~isolation. It was designed for use in~Blockly \cite{Blockly}, a~library for building visual programming editors.

JS-Interpreter supports only ES5\footnote{ECMAScript 5 or ECMAScript 2009} standard of JavaScript. This standard has no \texttt{class} keyword or block-scoped
variables with \texttt{let} keyword. We find ES5 way of defining classes using functions and prototypes confusing:
\begin{code}
// ES 5
function Person(name) {
    this.name = name;  
}

Person.prototype.greet = function () {
    console.log('Hello my name is ' + this.name);
}

// ES 6
class Person {
    constructor(name) {
        this.name = name
    }

    greet() {
        console.log('Hello my name is ' + this.name);
    }
}
\end{code}

We could add support for ES6 features to the~JS-Interpreter and use it for our project. However, as in~the~case of Brython we decided to rather put the~effort into
creating a~new solution than~putting a~lot of effort an~existing one.

\section{Text format}
Before we dive into the~language itself, let us first discuss the~underlying representation of the~language. We could represent programs either
in a \emph{binary format} or a \emph{textual format}. Using a binary format, we could map images to numbers and considering that our ``alphabet of images''
would be at most 256 characters long, we could use one byte per image. However, it is much simpler to develop the language with a textual format because
when writing tests and debugging we can work with printable characters rather than seemingly random numbers. Therefore, we decided to use a textual format.

We considered defining the~language in~terms of semicolon delimited tokens where each token could be substituted with a~single image.
Below is an~identity function defined in~both a~\emph{token-level} and a~\emph{character-level format}.
\begin{code}
// Token-level
FUNCTION; I;D; LPAR; A; RPAR; LCURLY;
RETURN; A; SEMICOLON;
RCURLY;

I;D; LPAR; 8; RPAR; SEMICOLON;

// Character-level
function ID(A) {
    return A;
}

ID(8);
\end{code}

A simpler option is to represent keywords as single characters in~the~character-level format. This makes the~image substitution trivial, as we can~substitute
every character. The~``keyword characters'' have to be unique among other characters. To illustrate this, we can~use a~Unicode letter `f' with a~hook $f$
to represent a~function and a~character $\hookleftarrow$ for return.
\begin{code}
$f$ ID(A) {
    $\hookleftarrow$ A;
}

ID(8);
\end{code}

The~above approach that uses Unicode characters has a~nice property regarding the~use of the~keywords in~a~string. According to the~ECMAScript specification,
Section 8.4 \cite{Ecmascript}: \emph{When a~String contains actual textual data, each element is considered to be a~single UTF-16 code unit.} This means that if we
only use characters that can~be encoded into a~single 16-bit code unit, all keywords will take space of a~single character in~a~string.

At first we used the~token-level format. It made the~implementation of the~parser harder because typical lexer responsibilities like processing
an~identifier had to be moved to the~syntactical analysis. Later on, we decided to switch to the~character-level format with keywords substituted for
Unicode characters. We changed the~format mostly because we realized that the~token-level format only introduced more complexity with no real benefits.

\section{Deciding on a~language type}
Since we have chosen to build our own language it is time to decide on what kind of language it should be and what features it should have. There are
multiple types of programming languages and we will briefly go through the~most known ones.

\subsection{Functional programming}
Functional programming treats computation as an~evaluation of mathematical functions and avoids changing state and mutable
data.\footnote{https://en.wikipedia.org/wiki/Functional\_programming}
This means that functional programming languages do not support imperative flow controls like loop statements.

Pure functions are very typical for functional programming. \emph{Pure functions} are functions whose outcome depends solely on their arguments
and create no side-effects (e.g., by changing a~global variable). Eliminating side effects can~make understanding a~program easier, which is one of the
key motivations for the~development of functional programming.

Additionally, functional programming often relies on recursion and functions being first-class entities. Being a~first-class entity means that functions can~appear
anywhere in~the~program where other first-class entities like numbers can, including arguments to other functions and their return values.

One of the~typical examples of a~functional programming language is Haskell. It was created in~1990, it is statically typed, supports type inference,
and lazy evaluation. Below is a~sample program in~Haskell finding the~edit distance between two strings from the~official docs.
\begin{code}
import Data.Array as Array

editDistance :: Eq a~=> [a] -> [a] -> Int
editDistance xs ys = table ! (m,n)
    where
    (m,n) = (length xs, length ys)
    x     = array (1,m) (zip [1..] xs)
    y     = array (1,n) (zip [1..] ys)
    
    table :: Array (Int,Int) Int
    table = array bnds [(ij, dist ij) | ij <- range bnds]
    bnds  = ((0,0),(m,n))
    
    dist (0,j) = j
    dist (i,0) = i
    dist (i,j) = minimum [table ! (i-1,j) + 1, table ! (i,j-1) + 1,
        if x ! i == y ! j
            then table ! (i-1,j-1)
            else 1 + table ! (i-1,j-1)]
\end{code}

\subsection{Object-oriented programing}
Object-oriented programming (OOP) is based around the~concept of ``objects'' which can~contain~data, in~the~form of fields (also known as attributes or properties),
and code, in~the~form of methods.\footnote{https://en.wikipedia.org/wiki/Object-oriented\_programming} Object's methods can~access and modify the~object they are
operating on using a~keyword like \texttt{this} or \texttt{self}. Programs then made out of objects that interact together.

The~most popular OOP languages are class-based (e.g., C++, Java, Python), meaning that objects are instances of classes, which also determine their types.
There are also alternatives, for example JavaScript is prototype-based.

OOP has a~few features that are common among most of the~object-oriented programming languages:
\begin{itemize}
    \item Inheritance or subclassing lets developers create a~new class by building upon an~existing one.
    \item Encapsulation allows hiding data~and methods of an~object from the~outside interference and misuse.
    \item Polymorphism allows calling code to be agnostic as to which class in~the~supported hierarchy it is operating on.
\end{itemize}

Below is an~example of polymorphism in~C++.
\begin{code}
class Animal {
public:
    virtual ~Base() { }
    virtual string make_sound() = 0;
};

class Dog: public Animal {
public:
    string make_sound() { 
        return "Woof, woof!"; 
    }
};

class Sheep: public Animal {
public:
    string make_sound() { 
        return "Baa, baa!"; 
    }
};

int main() {
    Animal *animal;
    Dog dog;
    Sheep sheep;

    // Store the~address of Dog.
    animal = &dog;

    // Call Dog's make_sound. Returns "Woof, woof!".
    animal->make_sound();
    // A~similar example with sheep. Returns "Baa, baa!".
    animal = &sheep;
    animal->make_sound();

    return 0;
}
\end{code}

\subsection{Esoteric programming languages}
An~esoteric programming language is designed to experiment with weird ideas, to be hard to program in, or as a~joke, rather than~for practical use. We can~not
describe esoteric languages in~general as they all use different ideas.

Probably the~most popular esoteric language is brainf***. It operates on an~array of memory cells, also referred to as the~tape. All of the~cells are initially
set to zero. There is a~single pointer, initially pointing to the~first memory cell. The~language has the~following commands:

\begin{figure}[!hbt]
    \centering
    \begin{tabular}{|c l |} 
    \hline
    $>$ & Move the~pointer to the~right \\
    \hline
    $<$ & Move the~pointer to the~left \\
    \hline
    $+$ & Increment the~memory cell under the~pointer \\
    \hline
    $-$ & Decrement the~memory cell under the~pointer \\
    \hline
    $.$ & Output the~character signified by the~cell at the~pointer \\
    \hline
    $,$ & Input a~character and store it in~the~cell at the~pointer \\
    \hline
    $[$ & Jump past the~matching ] if the~cell under the~pointer is 0 \\
    \hline
    $]$ & Jump back to the~matching [ if the~cell under the~pointer is nonzero \\
    \hline
    \end{tabular}
    \caption{a~list of brainf*** commands}
\end{figure}

Other examples of esoteric languages are:
\begin{itemize}
    \item Shakespeare, a~programming language whose source code resembles Shakespeare plays. 
    \item Befunge, a~language where the~code is layed out on a~two-dimensional grid of instructions, and execution can~proceed in~any direction of that grid.
\end{itemize}

\subsection{Conclusion}
There are also other programming language types apart from the~ones mentioned here, for example logic programming. We did not plan~on using
concepts from these languages and hence we did not list them. It would be great to combine a~language presented in~images with an~esoteric language
of some sort. However, we did not come up with an~interesting enough model and we did not want to blindly reuse an~existing one. in~the~end we settled on
a~more conservative approach.

Since our intended audience is typically experienced in~Python we decided to base our language around "Pythonic" features. We also think
the~class-based model of Python is more intuitive than~prototype-based one found for example in~JavaScript.

Even though the~author of this thesis likes statically-typed languages, Giflang is dynamically-typed. The~primary reason is that images take a~lot of
space on the~user screen and hence we tried to keep the~language concise. Adding typed declarations would probably make it less error-prone, but more verbose.

\section{Object model}
\label{chap3:object_model}
As mentioned in~the~end of the~previous section we base Giflang on Python. Everything is an~object in~Python. We will now explain~in~more detail what exactly
it means. an~object \texttt{A} can~have multiple related objects:
\begin{itemize}
    \item An~object \texttt{B} that denotes the~type of \texttt{A}. We call \texttt{B} a~\emph{class}\footnote{In the implementation we denote this \texttt{klass}
    instead of \texttt{class} because of a name collision with the keyword \texttt{class}.} of the~instance \texttt{A} and we say that \texttt{A}
    is an~instance of \texttt{B}.
    \item If \texttt{A} is a~class it can~have multiple \emph{base classes}. Let \texttt{B} denote one of the~base classes. We say that \texttt{A}
    \emph{derives from} \texttt{B}. 
\end{itemize}

Since classes are objects, they have to be instances of \emph{some} class. We call this ``class of classes'' a~\texttt{MetaClass}. Of course
\texttt{MetaClass} is also an~object and thus has to be an~instance of \emph{some} class. To break the~chain, \texttt{MetaClass} is an~instance of itself.
Figure \ref{fig:chap3:python_structure} describes the~Python's object structure in~more detail.

\begin{figure}[!hbt]
    \centering
    \tikzset{every picture/.style={line width=0.75pt}} %set default line width to 0.75pt        
    \trimbox{0cm .75cm 0cm 0cm}{
        \begin{tikzpicture}[x=0.75pt,y=0.75pt,yscale=-1,xscale=1]
        \draw   (13.61,159.25) -- (125.73,159.25) -- (125.73,200.19) -- (13.61,200.19) -- cycle ;
        \draw    (67.11,199.26) .. controls (87.27,268.92) and (-12.63,252.81) .. (12.38,202.52) ;
        \draw [shift={(13.61,200.19)}, rotate = 479.36] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw   (14.54,80.16) -- (126.66,80.16) -- (126.66,121.1) -- (14.54,121.1) -- cycle ;
        \draw  [dash pattern={on 0.84pt off 2.51pt}]  (58.73,160.18) -- (58.73,129.48) -- (58.73,125.03) ;
        \draw [shift={(58.73,122.03)}, rotate = 450] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw    (82,122.03) -- (82,155.32) ;
        \draw [shift={(82,158.32)}, rotate = 270] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw   (163.41,159.25) -- (275.54,159.25) -- (275.54,200.19) -- (163.41,200.19) -- cycle ;
        \draw    (162.95,179.72) -- (129.66,179.72) ;
        \draw [shift={(126.66,179.72)}, rotate = 360] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw  [dash pattern={on 0.84pt off 2.51pt}]  (216.45,159.25) -- (130.12,104.11) ;
        \draw [shift={(127.59,102.49)}, rotate = 392.57] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw   (313.22,149.02) -- (425.35,149.02) -- (425.35,209.5) -- (313.22,209.5) -- cycle ;
        \draw    (312.76,179.72) -- (279.47,179.72) ;
        \draw [shift={(276.47,179.72)}, rotate = 360] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw    (286.5,76.74) -- (223.15,76.74) ;
        \draw [shift={(220.15,76.74)}, rotate = 360] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw  [dash pattern={on 0.84pt off 2.51pt}]  (286.5,110.74) -- (223.15,110.74) ;
        \draw [shift={(220.15,110.74)}, rotate = 360] [fill={rgb, 255:red, 0; green, 0; blue, 0 }  ][line width=0.08]  [draw opacity=0] (10.72,-5.15) -- (0,0) -- (10.72,5.15) -- (7.12,0) -- cycle    ;
        \draw (69.67,179.72) node   [align=left] {{\fontfamily{pcr}\selectfont MetaClass}};
        \draw (70.6,100.63) node   [align=left] {{\fontfamily{pcr}\selectfont ObjectClass}};
        \draw (219.48,179.72) node   [align=left] {{\fontfamily{pcr}\selectfont UserClass}};
        \draw (370.87,179.72) node   [align=left] {{\fontfamily{pcr}\selectfont  UserClass}\\{\fontfamily{pcr}\selectfont instance}};
        \draw (351.87,75.72) node   [align=left] {{\fontfamily{pcr}\selectfont Instance of}};
        \draw (357.87,110.72) node   [align=left] {{\fontfamily{pcr}\selectfont Derives from}};
        \end{tikzpicture}
    }
    \caption{Object structure in~Python.}
	\label{fig:chap3:python_structure}
\end{figure}

\texttt{MetaClass} is actually called \texttt{type} in~Python. Below is an~interactive Python session that unveils the~internals of Python shown
in~Figure \ref{fig:chap3:python_structure}.
\begin{code}
>>> class X:
...     pass
... 
>>> type(X)
<class 'type'>
>>> x = X()
>>> type(x)
<class '__main__.X'>
>>> type(type(X))
<class 'type'>
>>> type(X).__bases__
(<class 'object'>,)
>>> type(type(X).__bases__[0])
<class 'type'>
\end{code}

We decided to reuse this model as the~base of Giflang. To make things simple we decided to not support multiple inheritance, though.

\subsection{Magic Methods}
Another Python feature we reused are magic methods. They provide a~very straightforward way
to overload operators. Magic methods are prefixed and suffixed with two underscores. Below is a~Python example that uses magic methods:
\begin{code}
class Number:
    # Defines a~constructor
    def __init__(self, val):
        self.val = val
    
    # Defines a~string representation
    def __str__(self):
        return str(self.val)

    # Overloads operator +
    def __add__(self, other):
        return Number(self.val + other.val)

    # Overloads a~call operator
    def __call__(self):
        return 'an~instance was called :)'

x = Number(2)
y = Number(3)
# Outputs 5. First __add__ is called and then __str__ on the~result.
print(x + y)
# Outputs 'an~instance was called :)'
print(x())
\end{code}

The~example shows how to create a~constructor, overload an~operator and create a~new instance of a~class by calling the~class with arguments for the~constructor.
\texttt{Number} class also defines a~\texttt{\_\_call\_\_} method so naturally we might expect the~call \texttt{Number(2)} to result in~the~string
`an~instance was called :)'. However, it is not the~case. Instead, \texttt{Number(2)} results in~a~new instance of the~\texttt{Number} class. The~author of this
thesis thinks that this is the~reason why magic method resolution in~Python starts from the~instance type rather than~from the~instance itself, as it is
the~case with regular methods.

Because the~call \texttt{Number(2)} does not result in~the~call of the~\texttt{\_\_call\_\_} function on the
\texttt{Number} type, it has to result in~a~\texttt{\_\_call\_\_} method further up the~chain. in~this case, \texttt{Number} is an~instance of the~\texttt{type},
or the~\texttt{MetaClass} class. Its \texttt{\_\_call\_\_} method is responsible for creating a~new instance and calling the~\texttt{\_\_init\_\_} method. Oversimplified,
the~\texttt{type}'s \texttt{\_\_call\_\_} function might look like this:
\begin{code}
class type:
    # Since classes are instances of the~'type' class, self
    # in~this method is a~class object. 
    def __call__(self, *args, **kwargs):
        obj = instantiate(self)
        obj.__init__(*args, **kwargs)
        return obj
\end{code}

in~reality, creating a~new instance is a~little more complicated. To give users more control over the~instantiation, another magic method (\texttt{\_\_new\_\_})
is called instead of a~made up function \texttt{instantiate} in~the~example above. We will not go into more details since we decided not to include the
\texttt{\_\_new\_\_} magic method into Giflang as it is mainly used only for a~few specific use cases (e.g., creating a~singleton).

Having a~keyword for introducing new variables, e.g., \texttt{var}, would mean~that magic methods could be resolved the~same way as normal methods because
it would allow us to distinguish between instantiation calls, e.g., \texttt{Number(3)}, and regular method call. However,
we like the~simplicity of creating a~new instance via~calling a~class and hence used this approach in~Giflang as well.

\section{Basic types}
There are 7 basic types supported natively in~Giflang: \texttt{Object}, \texttt{Number}, \texttt{Boolean}, \texttt{None}, \texttt{Function},
\texttt{Array} and \texttt{String}.

Since the~interpreter is written in~JavaScript, we based the~types mostly around JavaScript types; for example \texttt{Number} is a~wrapper of the~JavaScript
\texttt{number}. Because of this, it has a~standard double-precision 64-bit floating point format (IEEE 754). Booleans \texttt{True} and \texttt{False}
and also \texttt{None} are all represented using a~single image. in~the~language itself they are just global variables, similar to Python.

Functions are first-class entities, which means that users can~create anonymous functions, pass callbacks, etc. Closures are also supported.
Arrays and strings are basically wrappers around JavaScript arrays and strings as it was the~simplest approach. There are no primitive types and everything is
an~\texttt{Object}. Of course, it would be more effective to implement the~\texttt{Number} as a~primitive type, but as stated earlier, performance was not one
of the~main~objectives and implementing the~\texttt{Number} as an~\texttt{Object} was easier.

As with any other language, there needs to be a~support for basic conversions. We could create explicit conversion functions, but since Giflang is dynamically typed,
we implemented conversions in~the~constructors of the~target types. For example, calling \texttt{Number('1')} should result in~a~conversion from a~\texttt{String} to
a~\texttt{Number}.

\section{Control flow constructs}
Giflang supports the~following control flow constructs: \texttt{if}, \texttt{while} and \texttt{for}. The author of this thesis is a~fan~of the~C-like \texttt{for}
and thus used it in~Giflang as well. When evaluating a~condition, a~\texttt{\_\_bool\_\_} magic method is called on the~resulting object to convert it to
a~\texttt{boolean}. By default, everything except for \texttt{0}, \texttt{None}, \texttt{False}, an~empty string and an~empty array, evaluates to \texttt{True}.

\section{Standard library}
Giflang's standard library is quite modest. There is only support for the~very basics: \texttt{array}, \texttt{string}, \texttt{I/O}. As mentioned earlier, arrays
and strings wrap standard JavaScript arrays and strings. The~I/O is provided in~the~form of three functions:
\begin{itemize}
    \item \texttt{input} function reads a~single line
    \item \texttt{println} calls \texttt{\_\_str\_\_} magic method on its arguments to stringify them and outputs them on a~single line, space separated and
    with a~newline at the~end
    \item \texttt{print} has the~same semantics as \texttt{println}, but omits the~newline at the~end
\end{itemize}

We did not extend the~library further since arrays, strings and I/O are the~necessary building blocks that should suffice for most of the~basic use-cases.
Extending the~library further would not be hard, but we decided to invest the~time into different areas of the~project.